---
title: "**Exploratory Data Analysis**"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r,echo = FALSE,message= FALSE, warning = FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(stringr)
library(readxl)
```

### **Data Cleanning**

#### Voting result and Covid-19
```{r}
covid_19 = read.csv("data/New_York_State_Statewide_COVID-19_Testing__Archived__20241205.csv")
covid_19_NY = filter(covid_19,
                     Test.Date == "11/03/2020" &
                     Geography == "COUNTY")
covid_19_fatality = read.csv("data/New_York_State_Statewide_COVID-19_Fatalities_by_County_20241205.csv")
covid_19_fatality_NY = filter(covid_19_fatality,
                     Report.Date == "11/03/2020")
election_result = read.csv("data/election_result_NY_2020.csv")
election_covid = left_join(covid_19_NY, election_result,by = "County")
election_covid = left_join(election_covid, covid_19_fatality_NY, by = "County")
```

#### Socioeconomic indicator---poverty
```{r}
#Load dataset
poverty_level <- read.csv("data/2000_socioeconomic_indicator_poverty.csv") %>%
  janitor::clean_names()

#Extract the first column and use it as a dataframe
first_column <- data.frame(poverty_level$label_grouping)%>%
  filter(str_ends(poverty_level$label_grouping, "New York"))

#keeping all estimate as a seprate dataframe
df <- poverty_level%>%
  filter(str_ends(label_grouping,"Estimate"))

#Binding those two into one dataframe
poverty_county_level <- cbind(first_column, df) %>%
  mutate(
    total = as.numeric(gsub(",", "", total)),
    total_income_in_the_past_12_months_at_or_above_the_poverty_level = as.numeric(gsub(",", "", total_income_in_the_past_12_months_at_or_above_the_poverty_level)),
    total_income_in_the_past_12_months_below_poverty_level = as.numeric(gsub(",", "", total_income_in_the_past_12_months_below_poverty_level)),
    above_poverty_percentage = (total_income_in_the_past_12_months_at_or_above_the_poverty_level / total) * 100,
    below_poverty_percentage = (total_income_in_the_past_12_months_below_poverty_level / total) * 100
  ) %>%
  rename(
    county = poverty_level.label_grouping
  ) %>%
  mutate(
    county = sub(" County, New York$", "", county)
  ) %>%
  select(-label_grouping)%>%
  slice(-1) 
```

#### 2020 voting demographic---age
```{r, message= FALSE,warning = FALSE}
#load the dataset
voting_age <- read_csv("data/2000_voting_demographic_age.csv")%>%
  janitor::clean_names()

#Extract the first column and use it as a dataframe
first_column_age <- data.frame(voting_age$label_grouping)%>%
  filter(str_ends(voting_age$label_grouping, "New York"))%>%
   rename(
    county = voting_age.label_grouping
  ) %>%
    mutate(
    county = sub(" County, New York$", "", county))

#keeping all estimate as a seprate dataframe

df_age <- voting_age%>%
  filter(str_ends(label_grouping,"Estimate"))%>%
  mutate(
    total_18_to_29_years = as.numeric(gsub(",", "",total_18_to_29_years)),
    total_30_to_44_years = as.numeric(gsub(",", "",total_30_to_44_years)),
    total_45_to_64_years = as.numeric(gsub(",", "",total_45_to_64_years)),
    total_65_years_and_over = as.numeric(gsub(",", "",total_65_years_and_over)),
    total = as.numeric(gsub(",", "",total)))%>%
  mutate(
    below_30_percent = (total_18_to_29_years)/total*100,
    above_30_percent = (total_30_to_44_years+total_45_to_64_years+total_65_years_and_over)/total*100,
    above_65_percent = total_65_years_and_over/total*100)

#Binding those two into one dataframe
voting_age_county_level <- cbind(first_column_age, df_age)%>%
  slice(-1)
```

### **hospital counts in New York States**
```{r}
#load the dataset
health_facility_count <- read_excel("data/2020_health_facility_count.xlsx")

#create barchart of hospital distribution
ggplot(health_facility_count, aes(x = reorder(county, count), y = count)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Healthcare Infrastructure Counts by County",
       x = "County",
       y = "Count of Hospitals") +
  theme(axis.text.y = element_text(size = 6),
        axis.text.x = element_text(size = 10))
```

### **hospital counts in New York States Republician vs. Democratic**
```{r}
#Getting election results in binary version
election_result <- election_result%>%
  janitor::clean_names()%>%
  mutate(
    result = if_else(democratic_percent > republician_percent, "democratic", "republic"))%>%
  select(county,result)

#merge tables to get hospital count classified by rep vs. demo
election_result_hospital_count <- merge(election_result, health_facility_count, by = "county", all.x = TRUE)

#create barchart for hospital count based on election results
count_result <- aggregate(count ~ result, data = election_result_hospital_count, FUN = sum, na.rm = TRUE)

ggplot(count_result, aes(x = reorder(result, count), y = count, fill = result)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = c("democratic" = "blue", "republic" = "red")) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Hospital Counts by County --- democratic vs. republic",
       x = "",
       y = "Count of Healthcare Infrastructure",
       fill = "Result")
```

### **voting age distribution in New York State**
```{r}
#load the dataset
voting_age_county_level_clean <- read_excel("data/voting_age_county_level_clean_v2.0.xlsx")%>%
  select(county,total_18_to_29_years,total_30_to_44_years,total_45_to_64_years,total_65_years_and_over)

#create long dataframe
long_data <- voting_age_county_level_clean %>%
  pivot_longer(
    cols = starts_with("total_"), 
    names_to = "age_group",       
    values_to = "population")

#create barchart
ggplot(long_data, aes(x = reorder(county, population), y = population, fill = age_group)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  coord_flip() +  
  facet_wrap(~age_group, scales = "free_x") + 
  theme_minimal() +
  labs(
    title = "Population Distribution by Age Group and County",
    x = "County",
    y = "Population Count"
  ) +
  theme(
    strip.text = element_text(size = 7, face = "bold"), 
    axis.text.y = element_text(size = 6),  
    axis.text.x = element_text(size = 10),  
    plot.title = element_text(size = 14, face = "bold")
  )
```

### **voting age distribution in New York State Republician vs. Democratic**
```{r}
#merge tables to get age_distribution classified by rep vs. demo
election_result_voting_age_distribution <- merge(election_result, voting_age_county_level_clean, by = "county", all.x = TRUE)

#merge dataset for age distribution for democratic vs. republican
voting_age_result_1 <- aggregate(total_18_to_29_years ~ result, data = election_result_voting_age_distribution, FUN = sum, na.rm = TRUE)

voting_age_result_2 <- aggregate(total_30_to_44_years ~ result, data = election_result_voting_age_distribution, FUN = sum, na.rm = TRUE)

voting_age_result_3 <- aggregate(total_45_to_64_years ~ result, data = election_result_voting_age_distribution, FUN = sum, na.rm = TRUE)

voting_age_result_4 <- aggregate(total_65_years_and_over ~ result, data = election_result_voting_age_distribution, FUN = sum, na.rm = TRUE)

voting_age_result_df <- Reduce(function(x, y) merge(x, y, by = "result", all.x = TRUE), 
                               list(voting_age_result_1, voting_age_result_2, voting_age_result_3, voting_age_result_4))

#create barchart for voting age distribution based on election results
voting_age_long <- pivot_longer(
  voting_age_result_df,
  cols = starts_with("total_"),
  names_to = "age_group",
  values_to = "population"
)

# Create the bar chart
ggplot(voting_age_long, aes(x = result, y = population, fill = age_group)) +
  geom_bar(stat = "identity", position = "stack", color = "black") +
  theme_minimal() +
  labs(
    title = "Population Distribution by Voting Age and Result",
    x = "Result",
    y = "Population Count",
    fill = "Age Group"
  ) +
  scale_fill_brewer(palette = "Set3") + 
  theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )
```

### **poverty distribution in New York State**
```{r}
ggplot(poverty_county_level, aes(x = reorder(county, below_poverty_percentage), y = below_poverty_percentage)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    title = "Poverty Rate by County",
    x = "County",
    y = "Poverty Rate (%)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### **COVID-19 positive rate distribution in New York State**
```{r}
#Load the dataset and merge with election results based on county
voting_covid_detail <- read_csv("data/voting_covid_detail.csv")%>%
  janitor::clean_names()%>%
  left_join(election_result,by="county")%>%
  select(county,result,test_positive)

#create barchart of test_positive distribution
ggplot(voting_covid_detail, aes(x = reorder(county, test_positive), y = test_positive)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Healthcare Infrastructure Counts by County",
       x = "County",
       y = "Count of Hospitals") +
  theme(axis.text.y = element_text(size = 6),
        axis.text.x = element_text(size = 10))
```


### **COVID-19 positive rate distribution in New York State - Republician vs. Democratic**

```{r,message= FALSE,warning = FALSE}
#Create dataframe of average test_positive rate in county --- Republician vs. Democratic
test_positive_result <- aggregate(test_positive ~ result, data = voting_covid_detail, FUN = mean, na.rm = TRUE)

# Create the bar chart
ggplot(test_positive_result, aes(x = reorder(result, test_positive), y = test_positive, fill = result)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = c("democratic" = "blue", "republic" = "red")) +
   ylim(0, max(test_positive_result$test_positive) * 1.4) +
  coord_flip()+
  theme_minimal() +
  labs(
    title = "Average Test-Positive Rate by Election Result",
    x = "",
    y = "Average Test-Positive Rate",
    fill = "Result"
  )
```

