---
title: "Model of Republician voting rate"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r, echo = FALSE,message= FALSE, warning = FALSE}
library(tidyverse)
library(readxl)
library(modelr)
library(mgcv)
library(broom)
```

```{r, echo = FALSE}
election_covid = read.csv("data/voting_covid_detail.csv")
age = read_excel("data/voting_age_county_level_clean_v2.0.xlsx")
poverty = read_excel("data/poverty_county_level__clean_v2.0.xlsx")
hospitals = read_excel("data/2020_health_facility_count.xlsx")
names(election_covid)[names(election_covid)=="County"] <- "county"
```

```{r, echo = FALSE}
final_data = list(election_covid,age,poverty,hospitals)
final_data = final_data %>% reduce(left_join, by = "county")
```


### Linear Model of ELection Results

```{r,message= FALSE, warning = FALSE}
model1 = lm(republician_percent~below_poverty_percentage+below_30_percent+above_65_percent+count, data = final_data)

model1_data <- final_data |>
  add_predictions(model1) |>
  add_residuals(model1)

ggplot(model1_data, aes(x = pred, y = resid)) +
  geom_point(alpha = 0.2) +
  labs(title = "Residual against predicted value regardless of COVID-19",
       x = "Predicted value",
       y = "residual value") +
  geom_smooth()
```
*Formula: republican_percent∼below_poverty_percentage+below_30_percent+above_65_percent+count

Description: This baseline linear regression model evaluates the relationship between Republican voting percentages and key demographic factors at the county level, excluding any variables related to COVID-19. These predictors include poverty levels, the proportion of the voting- population below 30 years old, the proportion of the voting- populationabove 65 years old, and total population count in the county.

Residual Analysis: A scatter plot of residuals versus predictions reveals potential patterns in prediction errors, indicating whether the model assumptions hold and whether additional factors might improve predictive power.

Null Hypothesis: Each coefficient of the independent variables (below_poverty_percentage, below_30_percent,above_65_percent, count) is equal to zero (H0:βi=0). This implies that these demographic factors have no effect on Republican voting percentages.


### Linear Model of ELection Results including COVID-19
```{r,message= FALSE, warning = FALSE}
model2 = lm(republician_percent~below_poverty_percentage+below_30_percent+above_65_percent+count+Test_Positive, data = final_data)

model2_data <- final_data |>
  add_predictions(model2) |>
  add_residuals(model2)

ggplot(model2_data, aes(x = pred, y = resid)) +
  geom_point(alpha = 0.2) +
  labs(title = "Residual against predicted value considering COVID-19",
       x = "Predict value",
       y = "residuals")+
  geom_smooth()
```
Formula: republican_percent∼below_poverty_percentage+below_30_percent+above_65_percent+count+Test_Positive

Description: This model extends reduced Model by including the variable (Test_Positive), representing COVID-19 test positivity rates at the county level. This addition captures the potential influence of the pandemic on voting behaviors.

Residual Analysis: As with reduced model, residuals are analyzed against predictions to assess the model's accuracy and the effect of COVID-19 variable.

Null Hypothesis: Each coefficient of the independent variables, including Test_Positive, is equal to zero (H0:βi=0). Specifically, for the COVID-19 variable, H0: βTest_Positive=0, implying that the test postivity rate has no significant impact on Republican voting percentages.



### Comparasion between reduced model and full model
```{r}
compare = crossv_mc(final_data, 100) |> mutate(train = map(train, as_tibble),
                                      test = map(test, as_tibble))

compare = compare |> mutate(
    model1 = map(train, \(df) lm(republician_percent~below_poverty_percentage+below_30_percent+above_65_percent+count, data = df)),
    model2 = map(train, \(df) lm(republician_percent~below_poverty_percentage+below_30_percent+above_65_percent+count+Test_Positive, data = df)),
    rmse_model1 = 
      map2_dbl(model1, test, \(mod, df) rmse(model = mod, data = df)),
    rmse_model2 = 
      map2_dbl(model2, test, \(mod, df) rmse(model = mod, data = df))
)

compare |> 
  select(starts_with("rmse")) |> 
  pivot_longer(
    everything(),
    names_to = "model", 
    values_to = "rmse",
    names_prefix = "rmse_") |> 
  mutate(model = fct_inorder(model)) |> 
  ggplot(aes(x = model, y = rmse)) + geom_violin()
```
Methodology: A Monte Carlo cross-validation procedure with 100 iterations is used to compare reduced model and full model based on their root mean squared error (RMSE) performance. The dataset is repeatedly split into training and testing sets, and both models are fitted on training data and evaluated on testing data.

Visualization: A violin plot of RMSE values for the two models is generated to compare their predictive accuracy. The model with lower and more consistent RMSE values is considered superior in predictive performance. By incorporating demographic and pandemic-related factors, these models collectively aim to understand the dynamics influencing Republican voting rates and assess the impact of COVID-19  on the 2020 election results.

Null Hypothesis: There is no significant difference in predictive performance between Model 1 and Model 2 (H0:RMSEModel1=RMSEModel2). This implies that including the COVID-19 test positivity rate does not improve the model's accuracy in predicting Republican voting percentages. These hypotheses are tested to evaluate the significance of each variable and the overall model improvement when accounting for COVID-19. Statistical significance is typically assessed using p-values for individual coefficients and RMSE differences for model comparison.




